\bf Data Assimilation for Modeling and Calibrating of Disease-altered Potassium Channel Kinetics in Mouse Cardiomyocytes

### 2022-08-29 writing/draft.tex
Draft revised

I\textsubscript{Kslow2} has the same activation variable with I\textsubscript{Kslow1}, and the time-constant function for the inactivation $i_{\mathrm{Kslow2}}$ that contains the same steady-state function $i_{ss}$ in I\textsubscript{Kslow1}. As a result of this modeling strategy, mathematical equations of I\textsubscript{Kslow} are given as follows. 
\begin{align}
    &\mathrm{I}_{\mathrm{Kslow2}} = G_{\mathrm{Kslow2}}a_{\mathrm{Kslow2}}i_{\mathrm{Kslow2}}(V-E_{\mathrm{K}}) \\
    &a_{\mathrm{Kslow2}} = a_{\mathrm{Kslow1}} \\
    &\frac{di_{\mathrm{Kslow2}}}{dt} = \frac{i_{ss}-i_{\mathrm{Kslow2}}}{\tau_{i}^{(2)}} \\
    &\tau_{i}^{(3)} = p_{2} - p_{1}i_{ss}
\end{align}

@@

\section{Experimental Design}
More recently, we showed that preventing hybrid/complex N-glycosylation in mouse cardiomyocytes was sufficient to cause DCM using genetic ablation of the \textit{MGAT1} gene (MGAT1KO model), which encodes a critical glycosyltransferase, GlcNAcT1 \citep{ednie2019reduced2}. MGAT1KO mice developed DCM, deteriorated into heart failure, and 100\% died early, presumably from ventricular arrhythmias leading to sudden cardiac death. To further investigate the role of altered glycosylation in pathogenesis and disease progression of the heart, we conducted whole-cell patch-clamp experiments that showed reductions in N-glycosylation significantly impact electrical signaling in mouse cardiomyocytes \citep{ednie2019reduced}. To be specific, whole-cell I\textsubscript{K} traces were measured in left ventricular apex cardiomyocytes of ~14 week-old control (wild type; WT) and MGAT1KO mice, elicited by 4.5 s 10mV voltage steps (-30 - 50 mV) from holding potential of -70 mV. There were 31 and 30 observations for each WT and MGAT1KO.

@@

\subsection{Model Fitness}
We used the BFGS algorithm, one of the most popular nonlinear optimization methods, in the proposed model calibration routine because it supports box constraints.

box-constrained BFGS / goodness-of-fit measure / problem of rmse (not normalized) / problem of classical r2 / so, we chose a new r2 measure developed by UCLA guys.

We compared three nonlinear optimization algorithms satisfying box constraints: BFGS, sequential quadratic programming (SQP), and active set methods. \textbf{Fig~\ref{fig:model_fitness}A} and \textbf{C} shows values of the objective function (sum of RMSEs) of the three optimization algorithms for WT and MGAT1KO, respectively. The three algorithms barely showed differences in the objective values. We compared the actual model predictions and experimental potassium I\textsubscript{K} traces in \textbf{Figure~\ref{fig:model_fitness}B} and \textbf{D} to validate the model calibration results further. Representative data were selected randomly for each group and plotted at -30 mV and 50 mV voltage steps to illustrate discrepancies at low and high voltages. Sum of RMSEs values are lower in MGAT1KO than WT in general, because magnitudes are smaller for the disease group. Calibration results of the BFGS algorithm will be used in analysis of kinetics modeling in the rest the sections. 
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/model_fitness_4half.pdf}
    \caption{Bar graphs of the sum of RMSEs of (A) WT and (C) MGAT1KO, and representative actual fitness between model outputs and experimental I\textsubscript{K} recordings (in red) at -30 mV and 50 mV voltage steps of (B) observation 26 in WT and (D) 4 in MGAT1KO.}
    \label{fig:model_fitness}
\end{figure}

\subsection{Kinetics Modeling} \label{s:results.kinetics}
The \textit{in-silico} modeling predicts that reduced N-glycosylations cause significant changes in channel kinetics and gating behavior. \textbf{Figure~\ref{fig:kinetics_ikto}} presents the \textit{in-silico} modeling results of showing kinetics variables to voltage relationships of K\textsubscript{v}4.2. There is a slight rightward shift in steady-state activation (SSA) values along the voltage axis (\textbf{Figure~\ref{fig:kinetics_ikto}A}), and steady-state inactivation (SSI) is reduced up to 0.2 at voltages lower 0 mV (\textbf{Figure~\ref{fig:kinetics_ikto}B}). Activation and inactivation time constants $\tau_{a}^{(1)}$ and $\tau_{i}^{(1)}$ to voltage relationships are shown in \textbf{Figure~\ref{fig:kinetics_ikto}C} and \textbf{D}. $\tau_{a}^{(1)}$ for MGAT1KO is higher in range of $>$ -10 mV than WT. $\tau_{i}^{(1)}$ for MGAT1KO is slightly shifted toward left. The shifts in SSA and SSI likely cause the reductions in I\textsubscript{Kto} magnitudes, and the increases time constants explain delays in the decaying phase observed in the \textit{in-vitro} experiments.
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/kinetics_ikto.pdf}
    \caption{Kinetics functions in I\textsubscript{Kto} and their relationships to voltage. (A) steady-state activation, (B) steady-state inactivation, (C) time-constant activation, and (D) time-constant inactivation.}
    \label{fig:kinetics_ikto}
\end{figure}

\textbf{Figure~\ref{fig:kinetics_rec}} shows kinetics modeling results of the channels producing the delayed rectifier currents, I\textsubscript{Kslow1} and I\textsubscript{Kslow2}, and steady-state current I\textsubscript{Kss}. The most prominent result is that the time constants of I\textsubscript{Kslow1} are estimated to be significantly slow in MGAT1KO $\sim$ 3.4 and 235.8 ms (\textbf{Figure~\ref{fig:kinetics_rec}D} and \textbf{E}). Similar to I\textsubscript{Kto}, there is a small rightward shift in SSA at voltaged $>$ -10 mV for MGAT1KO. SSI is also shifted $\sim$ 10 mV to the right. Slow time constants and increased SSI are likely to address the flattened shape of MGAT1KO currents as illustrated in \textbf{Figure~\ref{fig:model_fitness}D}). Time constant inactivation of I\textsubscript{Kslow2} (\textbf{Figure~\ref{fig:kinetics_rec}C}) and activation of I\textsubscript{Kss} (\textbf{Figure~\ref{fig:kinetics_rec}F}) do not show significant differences between the two groups.
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/kinetics_rec.pdf}
    \caption{Kinetics functions and their relationships to voltage of I\textsubscript{Kslow1} (A) steady-state activation, (B) steady-state inactivation, (D) time-constant activation, (E) time-constant inactivation, (C) I\textsubscript{Kslow2} time-constant inactivation, and (F) I\textsubscript{Kss} time-constant activation.}
    \label{fig:kinetics_rec}
\end{figure}

\subsection{Parameter Distributions Between WT and MGAT1KO} \label{s:results.distribution}
We estimated distributions of the calibrated kinetic parameters to analyze the impacts of reduced N-glycosylation further. \textbf{Figure~\ref{fig:hist_pkto}} shows the distributional differences of the kinetic parameters in the K\textsubscript{4.2} model between the two group. A notable fact about the figure is that MGAT1KO shows small variances for Parameters 1, 5, and 7 in transition $\alpha_{a}$ and $\beta_{a}$, which control kinetics activation of I\textsubscript{Kto}. Reduced glycosylation affects activation gating that plays a vital role in I\textsubscript{Kto}, as it is characterized by a high peak at the early depolarization and rapid inactivation. In addition, MGAT1KO causes a leftward shift in the distribution of the maximum conductance $G_{\mathrm{Kto}}$, resulting in smaller conduction than WT in general. It is likely due to the observed current reductions in MGAT1KO.
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/hist_pkto.pdf}
    \caption{Distribution estimations of calibration parameters in the K\textsubscript{Kto} model.}
    \label{fig:hist_pkto}
\end{figure}

@@

\subsection{Low-dimensional Embedding and Visualization}
We aggregated calibration results into one to analyze the parameters in a collective way. It resulted in a $61\times20$ matrix (i.e., 31 observations in WT and 30 in MGAT1KO, and there are 20 calibration parameters). Then we applied t-distributed stochastic neighbor (t-SNE) embedding to encapsulate the high-dimensional data into three-dimensional space so that we could visualize each data point. \textbf{Figure~\ref{fig:cluster}} presents the visualization of the t-SNE embedding of the calibration results. It shows not only clear differences between the two groups but also variances of cells (i.e., data points) within the groups and across the entire dataset.
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/cluster.pdf}
    \caption{t-SNE embedding of calibration parameters.}
    \label{fig:cluster}
\end{figure}

>>>>

I\textsubscript{Kslow2} has the same activation variable with I\textsubscript{Kslow1}, and the time-constant function of the inactivation $i_{\mathrm{Kslow2}}$ that contains the same steady-state function $i_{ss}$ in I\textsubscript{Kslow1}. As a result of this modeling strategy, mathematical equations of I\textsubscript{Kslow} are given as follows. 
\begin{align}
    &\mathrm{I}_{\mathrm{Kslow2}} = G_{\mathrm{Kslow2}}a_{\mathrm{Kslow2}}i_{\mathrm{Kslow2}}(V-E_{\mathrm{K}}) \\
    &a_{\mathrm{Kslow2}} = a_{\mathrm{Kslow1}} \\
    &\frac{di_{\mathrm{Kslow2}}}{dt} = \frac{i_{ss}-i_{\mathrm{Kslow2}}}{\tau_{i}^{(2)}} \\
    &\tau_{i}^{(3)} = p_{2} - p_{1}i_{ss}
\end{align}

@@

\section{Experimental Design}
More recently, we showed that preventing hybrid/complex N-glycosylation in mouse cardiomyocytes was sufficient to cause DCM using genetic ablation of the \textit{MGAT1} gene (MGAT1KO model), which encodes a critical glycosyltransferase, GlcNAcT1 \citep{ednie2019reduced2}. MGAT1KO mice developed DCM, deteriorated into heart failure, and 100\% died early, presumably from ventricular arrhythmias leading to sudden cardiac death. To further investigate the role of altered glycosylation in pathogenesis and disease progression of the heart, we conducted whole-cell patch-clamp experiments that showed reductions in N-glycosylation significantly impact electrical signaling in mouse cardiomyocytes \citep{ednie2019reduced}. To be specific, whole-cell I\textsubscript{K} traces were measured in left ventricular apex cardiomyocytes of ~14 week-old control (wild type; WT) and MGAT1KO mice, elicited by 4.5 s 10mV voltage steps (-30 - 50 mV) from holding potential of -70 mV. There were 32 observations for each WT and MGAT1KO.

@@

\subsection{Model Fitness}
We used the BFGS algorithm, one of the most popular nonlinear optimization methods, in the proposed model calibration routine because it supports box constraints. A separate accuracy metric is used to measure the goodness-of-fit and potential prediction power of the calibration results rather than utilizing the final objective function value of the calibration routine to report the fitness. The objective function, the sum of RMSEs over voltage steps, does not provide a straightforward interpretation that makes it difficult to measure the performance of the models. $R^{2}$, defined in Equation~\ref{eq:r2}, is the most common prediction accuracy metric ranging from 0 to 1 for linear regression models that represents the proportion of variation in data that is explained by the model. However, this interpretation is not applicable, and its value can be negative for nonlinear models because the variance decomposition in Equation~\ref{eq:var_decomp} does not hold anymore.
\begin{align}
    R^{2} &= 1 - \frac{\sum_{k=1}^{n}(y^{\mathcal{D}}(k)-y^{\mathcal{M}}(k|\theta))^2}{\sum_{k=1}^{n}(y^{\mathcal{D}}(k)-\bar{y})^2} \label{eq:r2} \\
    \sum_{k=1}^{n}(y^{\mathcal{D}}(k)-\bar{y})^2 &= \sum_{k=1}^{n}(y^{\mathcal{M}}(k|\theta))-\bar{y})^2+\sum_{k=1}^{n}(y^{\mathcal{D}}(k)-y^{\mathcal{M}}(k|\theta))^2 \label{eq:var_decomp}
\end{align}

Therefore, a pseudo-R-squared measure, proposed in \citet{li2019prediction} was used to evaluate the model fitness in this study. This nonlinear $R^{2}$ measure is based on the concept of linear correction of prediction function, allowing the variance decomposition for nonlinear functions, which guarantees straightforward interpretation and normalized scores between 0 and 1 as in the classical $R^{2}$. \textbf{Fig~\ref{fig:model_fitness}A} and \textbf{C} present averages of $R^{2}$ values over voltage steps from -30 to 50 mV by 10 mV step size for WT and MGAT1KO, respectively. In most cases, more than 80 \% of variances on average in multiple current recordings are explained by models (WT: $0.9624 \pm 0.0051$; MGAT1KO: $0.8938 \pm 0.0077$). The model fitness measures in MGAT1KO are less than WT in general. It is expected that the small current magnitude of MGAT1KO challenges the calibration process. We compared the actual model predictions and experimental I\textsubscript{K} traces in \textbf{Figure~\ref{fig:model_fitness}B} and \textbf{D} to validate the model calibration results further. As illustrated in the comparison plot of MGAT1KO at -30 mV clamp voltage, small and noisy current trace results in decreased $R^{2}$, 0.8335. 
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/model_fitness_4half.pdf}
    \caption{Bar graphs of the averages of nonlinear $R^{2}$s for nine voltage steps from -30 mV to 50 mV by the 10 mV step size for (A) WT and (C) MGAT1KO. Exemplary actual fitness between model predictions (in green) and experimental I\textsubscript{K} recordings (in pink) at -30 mV and 50 mV voltage steps of (B) cell 6 in WT and (D) 12 in MGAT1KO.}
    \label{fig:model_fitness}
\end{figure}

\subsection{Kinetics Modeling} \label{s:results.kinetics}
The \textit{in-silico} modeling predicts that reduced N-glycosylations cause significant changes in channel kinetics and gating behavior. \textbf{Figure~\ref{fig:kinetics_ikto}} presents the \textit{in-silico} modeling results showing kinetics variables to voltage relationships of K\textsubscript{v}4.2. There is a slight rightward shift in steady-state activation (SSA) values along the voltage axis (\textbf{Figure~\ref{fig:kinetics_ikto}A}), and steady-state inactivation (SSI) is reduced up to 0.25 at voltages lower -10 mV (see \textbf{Figure~\ref{fig:kinetics_ikto}B}). Activation and inactivation time constants ($\tau_{a}^{(1)}$ and $\tau_{i}^{(1)}$) to voltage relationships are shown in \textbf{Figure~\ref{fig:kinetics_ikto}C} and \textbf{D}. $\tau_{a}^{(1)}$ for MGAT1KO is greater than WT in range of $>$ -20 mV. $\tau_{i}^{(1)}$ for MGAT1KO shifted toward left. The shifts in SSA and SSI likely cause the reductions in I\textsubscript{Kto} magnitudes, and the increased time constants explain delays in the decaying phase observed in the \textit{in-vitro} experiments.
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/kinetics_ikto.pdf}
    \caption{Kinetics functions and their relationships to voltage of I\textsubscript{Kto} (A) steady-state activation, (B) steady-state inactivation, (C) activation time constant, and (D) inactivation time constant.}
    \label{fig:kinetics_ikto}
\end{figure}

\textbf{Figure~\ref{fig:kinetics_rec}} shows kinetics modeling results of the channels producing the delayed rectifier currents, I\textsubscript{Kslow1} and I\textsubscript{Kslow2}, and steady-state current I\textsubscript{Kss}. The most prominent result is that time constants of activation and inactivation in I\textsubscript{Kslow1} ($\tau_{a}^{(2)}$ and $\tau_{i}^{(2)}$) are estimated to be significantly slow in MGAT1KO, i.e., $\sim$ 3.5 and 348.9 ms, respectively (see \textbf{Figure~\ref{fig:kinetics_rec}D} and \textbf{E}). Similar to I\textsubscript{Kto}, there is a small rightward shift in SSA at voltaged $>$ -10 mV for MGAT1KO. SSI is also shifted $\sim$ 10 mV to the right. The slow time constants and increased SSI are likely to address the flattened and reduced currents of MGAT1KO. Time constant inactivation of I\textsubscript{Kslow2} ($\tau_{i}^{(3)}$) is also increased (see \textbf{Figure~\ref{fig:kinetics_rec}C}), but activation of I\textsubscript{Kss} ($\tau_{a}^{(4)}$) do not show significant differences between the two groups (see \textbf{Figure~\ref{fig:kinetics_rec}F}).
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/kinetics_rec.pdf}
    \caption{Kinetics functions and their relationships to voltage of I\textsubscript{Kslow1} (A) steady-state activation, (B) steady-state inactivation, (C) activation time constant, (D) inactivation time constant, (E) I\textsubscript{Kslow2} inactivation time constant, and (F) I\textsubscript{Kss} activation time constant.}
    \label{fig:kinetics_rec}
\end{figure}

\subsection{Parameter Distributions Between WT and MGAT1KO} \label{s:results.distribution}
We estimated distributions of the calibrated kinetic parameters using histograms to analyze the impacts of reduced N-glycosylation further. \textbf{Figure~\ref{fig:hist_pkto}} shows the distributional differences of the kinetic parameters in the K\textsubscript{4.2} model between the two group. A notable fact about the figure is that MGAT1KO shows small variances for Parameters 1, 5, and 7 in transition $\alpha_{a}$ and $\beta_{a}$, which control kinetics activation of I\textsubscript{Kto}. Reduced glycosylation affects activation gating that plays a vital role in I\textsubscript{Kto}, as it is characterized by a high peak at the early depolarization and rapid inactivation. In addition, MGAT1KO causes a leftward shift in the distribution of the maximum conductance $G_{\mathrm{Kto}}$, resulting in smaller conduction than WT in general. It is likely due to the observed current reductions in MGAT1KO.
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/hist_pkto.pdf}
    \caption{Distribution estimations of calibration parameters in the K\textsubscript{Kto} model.}
    \label{fig:hist_pkto}
\end{figure}

@@

\subsection{Low-dimensional Embedding and Visualization}
We aggregated calibration results into one to analyze the parameters in a collective way. It resulted in a $61\times20$ matrix (i.e., 31 observations in WT and 30 in MGAT1KO, and there are 20 calibration parameters). Then we applied t-distributed stochastic neighbor (t-SNE) embedding to encapsulate the high-dimensional data into three-dimensional space so that we could visualize each data point. \textbf{Figure~\ref{fig:embd_result}} presents the visualization of the t-SNE embedding of the calibration results. It shows not only clear differences between the two groups but also variances of cells (i.e., data points) within the groups and across the entire dataset.
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/embd_result.pdf}
    \caption{t-SNE embedding of calibration parameters.}
    \label{fig:embd_result}
\end{figure}
